% -*-latex-*-

\title{Scratching with All Your Fingers: Exploring Multi-Touch Programming in Scratch}

\author{Christopher Marten-Ellis Graves}
\prevdegrees{ S.B., Massachusetts Institute of Technology (2013)}
% If you wish to list your previous degrees on the cover page, use the 
% previous degrees command:
%       \prevdegrees{A.A., Harvard University (1985)}
% You can use the \\ command to list multiple previous degrees
%       \prevdegrees{B.S., University of California (1978) \\
%                    S.M., Massachusetts Institute of Technology (1981)}
\department{Department of Electrical Engineering and Computer Science}

% If the thesis is for two degrees simultaneously, list them both
% separated by \and like this:
% \degree{Doctor of Philosophy \and Master of Science}
\degree{Master of Engineering in Electrical Engineering and Computer Science}

% As of the 2007-08 academic year, valid degree months are September, 
% February, or June.  The default is June.
\degreemonth{June}
\degreeyear{2014}
\thesisdate{May 28, 2014}

%% By default, the thesis will be copyrighted to MIT.  If you need to copyright
%% the thesis to yourself, just specify the `vi' documentclass option.  If for
%% some reason you want to exactly specify the copyright notice text, you can
%% use the \copyrightnoticetext command.  


% If there is more than one supervisor, use the \supervisor command
% once for each.
\supervisor{Prof. Mitchel Resnick}{LEGO Papert Professor of Learning Research, MIT Media Lab}

% This is the department committee chairman, not the thesis committee
% chairman.  You should replace this with your Department's Committee
% Chairman.
\chairman{Prof. Albert R. Meyer}{Chairman, Masters of Engineering Thesis Committee}

% Make the titlepage based on the above information.  If you need
% something special and can't use the standard form, you can specify
% the exact text of the titlepage yourself.  Put it in a titlepage
% environment and leave blank lines where you want vertical space.
% The spaces will be adjusted to fill the entire page.  The dotted
% lines for the signatures are made with the \signature command.
\maketitle

% The abstractpage environment sets up everything on the page except
% the text itself.  The title and other header material are put at the
% top of the page, and the supervisors are listed at the bottom.  A
% new page is begun both before and after.  Of course, an abstract may
% be more than one page itself.  If you need more control over the
% format of the page, you can use the abstract environment, which puts
% the word "Abstract" at the beginning and single spaces its text.

%% You can either \input (*not* \include) your abstract file, or you can put
%% the text of the abstract directly between the \begin{abstractpage} and
%% \end{abstractpage} commands.

% First copy: start a new page, and save the page number.
\cleardoublepage
% Uncomment the next line if you do NOT want a page number on your
% abstract and acknowledgments pages.
% \pagestyle{empty}
\setcounter{savepage}{\thepage}
\begin{abstractpage}
\input{abstract}
\end{abstractpage}

% Additional copy: start a new page, and reset the page number.  This way,
% the second copy of the abstract is not counted as separate pages.
% Uncomment the next 6 lines if you need two copies of the abstract
% page.
% \setcounter{page}{\thesavepage}
% \begin{abstractpage}
% \input{abstract}
% \end{abstractpage}

\cleardoublepage

\section*{Acknowledgments}

I would first like to thank my advisor, Mitchel Resnick, for giving me the wonderful opportunity to be a part of the Lifelong Kindergarten Group. Thank you so much for your guidance and support!

I would also like to thank the extended Lifelong Kindergarten Group community for helping me throughout the year and for teaching me so much about so many things. Everyone I have met and gotten to know has been so welcoming, encouraging, and inspiring. Thank you all!

Special thanks to Champika Fernando for responding to my initial request to join the group a year ago and for serving as an extra mentor to me since then. I really appreciate all you have done for me. Thank you!

Next, I would like to thank Bob Irwin for reading through this entire thesis with me and helping me organize my abstract ideas into carefully organized words. I have learned a great deal through our meetings and I had fun, too. Thank you! 

I would also like to give special thanks to my partner, Julianna, for  putting up with me for the past $3\frac{1}{3}$ years and for providing me with endless encouragement. I love you!

Lastly, I would like to thank my family for providing me with perpetual support since the day I was born: my father, Stephen, for encouraging my curiosity and piquing my thirst for knowledge; my mother, Caroline, for inspiring me to never give up on following my dreams and to always think for myself; my godmother, Delmy, for showing me the values of working hard and being the best person I can be; and my sister, Carmen, for serving as my role model and personal trailblazer. I love you all very much!

 
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% -*-latex-*-
